### 2023-12-05 thesis/MSECE_thesis.tex
Last round of thesis edits.

\section{Data Sources}\label{sec:data_source}
There are four corpora used in this thesis. The datasets are primarily speech corpora intended either for automatic speech recognition research in the
domain of aviation communications, linguistics research in aviation English, or both. All data in all four corpora include both recordings and
transcriptions, audio and text, respectively. The transcriptions are used to pretrain and train the language models and tokenizers (described in
sections \ref{sec:language_modeling} and \ref{sec:tokenizers}).

@@

All data was processed in Python using primarily built-in functions/modules\footnote{The \lstinline|re| and \lstinline|xml| modules were also used
    although these are built into most Python distributions by default.}. The text data is extracted from the corpus transcripts by parsing from the
corpora-specific formats where necessary i.e.~Lisp in ATCC, XML in ATCO2 and ZCU CZ ATC, and plain text in ATCOSIM, the transcriber annotations
are removed where necessary\footnote{A copy of the original data is made, then the annotations are removed from the copy.}, and the resulting text
is written to a file in the standard corpus format (an ASCII text file with one line in the file corresponding to one sample from the corpus) to a
file.

\subsection{Handling Spelling Errors \& Inconsistencies}\label{sec:spelling_errors}
By the nature of the data labeling process, there are bound to be spelling errors introduced by the annotators during corpora creation. This gets
exacerbated by the aggregation of several corpora with varying controls for correcting spelling errors and inconsistencies. Nearly all of the errors
are recoverable/correctable given context. For example, ``possibility of \textbf{tornado's}'' should clearly have been ``possibility of
\textbf{tornadoes}'', changing the possessive form of ``tornado'' to the plural form. There are also instances of spelling conventions that are
consistent within corpora, but become inconsistent when the corpora are combined.

@@

% hesitations, cut off words, different kinds of filled pauses
The mapping system described above is also used to modify within corpora conventions to be consistent across corpora. Three areas are addressed here:
hesitations, cut-off or partial words, and filled pauses within speech. These areas are defined differently between the four corpora, which makes
unifying the transcripts difficult. Hesitations, for example, are defined by one corpus as any incomplete or non-english speech sounds and left
completely undefined by another (despite annotators making notes about hesitations). The way in which hesitations, pauses, and partial words are
transcribed differs between corpora as well. Two corpora prescribe special tokens to these phenomena such as ``[hes]'', ``[hesitation]'', ``<pause>'',
or similar, while the others transcribe the data as it sounds (i.e., ``uh'', ``eh'', or ``er'' for filled pauses and ``cir-'' or ``cir+'' for
instances that were cut-off mid pronunciation). In order to facilitate the unification of the transcripts for all four corpora, the conventions for
transcribing these phenomena are redefined below:

@@

The ATCC and ZCU CZ ATC corpora make up a significant portion of the data, ATCC alone makes up more than half (approximately 54\%) of the combined
data, and ATCC and ZCU CZ ATC together make up approximately 81\% of the data. The sequence lengths of the samples are fairly consistent across
corpora with at most about an 18\% percent difference between sequence lengths. The number of unique tokens does not scale additively to the
combined corpora, which is as expected. Considering that aviation English generally conforms to and follows ICAO phraseology standards we would expect
that many terms would be shared across corpora with the exception of geographic/location specific terms, such as city, landmark, airport, and airline
names, etc.~that typically vary across regions.

@@

Sci-Kit Learn's \cite{pedregosa_scikit-learn_2011} implementation of the LOF algorithm was used with a value of \(N = 35\) for \(N\) neighbors in the
K-Nearest Neighbors algorithm and Minkowski distance for distance metrics/calculations (in this case, since \(p=2\), Euclidean distance is effectively
used for distance calculations). A total of 47 samples were detected as outliers and removed for analysis and are summarized in Table
\ref{tab:outlier_stats}.

@@

Although all of the outliers have higher sequence lengths than the typical sequence lengths of the other samples in the data, going the through
the outlying samples manually and reviewing each reveals that, for the most part, the samples are typical albeit somewhat complex communications.
With the exception of four outliers, all of the samples labeled as outlying by the LOF algorithm were kept in the data. Examples of the outliers are
listed in Appendix A in Section \ref{sec:appendix_a}, including the sample removed.

\section{Language Modeling}\label{sec:language_modeling}
% bert, roberta
% n-gram
This section introduces the chosen language model architectures as well as their pretrained checkpoints, if one is used. Two transformer-based
architectures are used, BERT and RoBERTa, and statistical/machine learning-based approaches such as N-gram and word lattices.

The transformers are pre-trained using the same masked language modeling (MLM) objective presented in Devlin et al.~\cite{devlin_bert_2019} to train
BERT for bidirectional encoding representation and likewise in Liu et al.~\cite{liu_roberta_2019} for training RoBERTa. Given the results of the
RoBERTa pretraining and downstream task training thereafter, it was initially decided to eliminate the next sentence prediction (NSP) task as it was
used in BERT, however, their results suggest that the success of the MLM-only approach was due to the increase in batch and overall data
size~\cite{liu_roberta_2019}. Taking this into account, along with the very limited amount of data available in the domain of aviation an NSP
pretraining task was implemented where possible\footnote{Only about half of the data (only the data from ATCC, since it is continuous recordings of
    the airspace) is in a format that enables us to extract the turn taking phenomenon that is present in most air-traffic communications.} to improve
its pretraining results. MLM and NSP pretraining are explained more in-depth in Section \ref{sec:bert}, below.

@@

\textbf{Masked Language Modeling}. The ability of BERT to represent text in a bidirectional context is achieved through the Masked Language Modeling
(MLM) pretraining task~\cite{devlin_bert_2019}. MLM pretraining choses a certain percentage of token positions at random for prediction, then tokens
are replaced with either a ``mask'' token, a random token, or the original token according to preconfigured probabilities (BERT used 80\% for the mask
token, 10\% for random tokens, and 10\% for the original/unchanged token). The original token is then predicted with cross-entropy loss.

\textbf{Next Sentence Prediction}. While MLM conditions the model to represent tokens within sentences, NSP is meant to condition the model to
understand relationships between sentences/sequences to further support downstream tasks. Since BERTs training set consists primarily of literary
works and excerpts from Wikipedia, the  model simply predicts whether two sentences appear together (one after the other) in the original text,
making this a binary classification task as opposed to MLM which is a multi-class classification task. Instead of a single input sequence, the input
consists of two sequences separated by a ``separation'' token and beginning with a ``classification'' token, these are represented by
\lstinline|[SEP]| and \lstinline|[CLS]|, respectively in Devlin et al.~\cite{devlin_bert_2019}.

For symmetry between pretraining tasks, the special tokens specifying the beginning and end of input sequences are shared between tasks. For both MLM
and NSP objectives, input sequences always start with a classification token and end with a separation token (see Figure
\ref{fig:bert_mlm_input_example}). Likewise, NSP inputs always begin with a classification token and each input sequence is concatenated with a
separation token terminating each sequence (see Figure \ref{fig:bert_nsp_input_example} and equations (\ref{eq:bert_mlm_input_example}) and
(\ref{eq:bert_nsp_input_example})).

\begin{equation}\label{eq:bert_mlm_input_example}
    \mbox{[CLS], the, quick, brown, fox, [SEP]}
\end{equation}

@@

\begin{equation}\label{eq:bert_nsp_input_example}
    \mbox{[CLS], the, quick, brown, fox, [SEP], jumps, over, the, [SEP]}
\end{equation}

@@

The pretraining methodology introduced by RoBERTa eliminated the NSP pretraining task that was originally used in BERT and opted instead to increase
the quantity of training data along with the batch sizes in addition to training the model longer. Lastly, there is a small difference in the
construction of input sequences as compared to BERT; input sequences always begin with a classification token (\lstinline|[CLS]|) and individual
segments are separated by a separation token (\lstinline|[SEP]|). Rather than having sequences end with a separation token as well, they are
terminated with an end of sequence token (\lstinline|[EOS]|). See equation \ref{eq:roberta_input_example} modified from Liu et
al.~\cite{liu_roberta_2019}, below.

@@

\section{Tokenization Algorithms}\label{sec:tokenizers}
% present and explain the tokenization algorithms used (possibly which models they correspond to)
Tokenization is the process of separating or segmenting words in sentences into their component ``tokens''. The tokenized version of a sentence is
referred to as a sequence. The rules by which the tokens are created or segmented depend entirely on the algorithm used. The simplest and most
intuitive way to do this by creating tokens based on whitespace i.e., each ``word'' in the corpus is  treated as a token\footnote{The Huggingface
    Tokenizer library implements this as the \lstinline|WordLevel| tokenizer.}. While this is simple and works in theory, it does not take into
account the fact that not every word in a language will occur in a corpus. The inevitable possibility for a tokenizer to see a word that was not in
the original training corpus is known as the out-of-vocabulary (OOV) problem~\cite{wu_googles_2016}. The OOV problem has been addressed by several
different algorithms. For brevity's sake, only the tokenization algorithms used by the models in this work are defined and explained.

@@

In order to map individual tokens to numerical values or IDs, an index of the words that appear in a corpus must be created. The numerical values for
each token are derived from this index. For the WordLevel algorithm, this process is referred to as the training the tokenizer model (see algorithm
\ref{alg:wordlevel_training}).

@@

The tokenization of words with shared roots but different suffixes is shown in Table \ref{tab:wordlevel_tokenization_shared_roots_example}. Clearly
all three words are seen as independent and unique words by the tokenizer, since they are mapped to their own integer representations.

@@

\subsection{WordPiece}\label{sec:wordpiece}
% Wu et al. 2016; Schuster and Nakajima 2012
The WordPiece algorithm was developed by Schuster and Nakajima \cite{schuster_japanese_2012} as a word segmentation algorithm for language modeling in
Japanese and Korean voice search systems. It is further used by Wu et al.~\cite{wu_googles_2016} for neural machine translation tasks. The algorithm
demonstrated proficiency and increased performance for the models used in both tasks and effectively addressed the OOV problem.

One of the primary stipulations for this algorithm is effective and efficient handling of OOV words such that none are produced during normal
operation. To achieve this, during training, the vocabulary of the tokenizer is initialized to a basic set of characters; since WordPiece is designed
to be language agnostic, the initial vocabulary is specified as all basic unicode characters (for the chosen language) in addition to all ASCII
characters.

@@

Table \ref{tab:wordpiece_tokenization_example}, below, shows an example of a WordPiece tokenizer model output on a sentence that was seen during the
training procedure (trained on the same data as the WordLevel tokenizer in Section \ref{sec:wordlevel}). Since this sentence was in the training data,
the segmentation is identical to that of the WordLevel tokenizer in Table \ref{tab:wordlevel_tokenization_example}, however, tokenization of similar
words e.g.~words with shared roots or similar contractions are handled differently as in Table \ref{tab:wordpiece_tokenization_shared_roots_example}.
Since each word has a shared root word in the contractions (as well as an apostrophe separating the contraction), the tokenizer segments each word and
represents them as groups of small, more common (thus with higher probabilities of occurring) tokens. Additionally, the WordPiece algorithm
effectively handles OOV words. Take the OOV in Table \ref{tab:wordpiece_unk_word_similar} with a shared root and similar contraction as those in Table
\ref{tab:wordpiece_tokenization_shared_roots_example}, which appear in the training data. Since the base of the contraction, ``you'', occurs in the
training data as well as the apostrophe in the contraction, so the tokenizer segments those parts of the word as tokens. The last part of the word,
``ll'', does not appear at all in the training data, so it is broken down into its component characters which are represented as ``l'' (beginning of
a word) and ``\#\#l'' (segmented part of a word that should be concatenated to the previous token when they are recombined). Comparing this to the
results of the WordLevel tokenizer output on the same word in Table \ref{tab:wordlevel_unk_word}, the WordPiece algorithm sidesteps the OOV problem
for words similar to those in the trining corpus and, according to the output of the WordPiece tokenizer in Table \ref{tab:wordpiece_unk_word}, for
words unlike those in the training corpus.

@@

\subsection{Byte-Pair Encoding}\label{sec:bpe}
% Sennrich et al. 2016; Gage 1994
The Byte-Pair Encoding (BPE) tokenizer algorithm was adapted from a data compression algorithm by Gage 1994~\cite{gage_feb94_1994}. The data compression
algorithm works by replacing common pairs of bytes in data with an unused byte to reduce the overall size of the data. The tokenization algorithm
works under the same principle, merging frequent pairs of characters or tokens into one token~\cite{sennrich_neural_2016}.

@@

Now, testing the BPE algorithm on words that did not appear in the training data, it is clear that the BPE algorithm effectively deals with and
processes OOV words (tables \ref{tab:bpe_unk_similar} and \ref{tab:bpe_unk}), even when no subunits of the word appear in the training data (Table
\ref{tab:bpe_unk}).

@@

\textbf{Byte-Level Byte-Pair Encoding}. This is a subset of the BPE algorithm that is also commonly used for NLP tasks such as neural machine
translation, language modeling, and generative tasks (GPT-2 uses a byte-level BPE tokenizer~\cite{radford_language_2019}). As Huggingface's
\lstinline|tokenizers| library implements it, this is a preprocessing step for the tokenizer model that maps all bytes in a string/sentence to their
own unique and visible character. Functionally, byte-level BPE is nearly identical to standard BPE models except that the beginning of word symbol is
actually rendered and represented by a visible character in the tokenized sequence (see Table \ref{tab:byte_level_bpe_example}).

@@

\section{Discussion}\label{sec:discussion}
Considering the success of deep learning models with transformer-based architectures for language modeling and natural language processing (NLP)
tasks, more generally, in the conversational and literary English domains, it follows that transformer architectures should see widespread use in the
domain of aviation English for domain-specific NLP tasks, however, according to the literature, transformer-based language models have seen sparse
use. The most notable example of the use of a transformer deep neural network is the application of BERT for knowledge extraction from notice to
airmen (NOTAM) messages \cite{arnold_knowledge_2022}.

\subsection{Language Model Pretraining}\label{sec:lm_pretraining}
A standard masked language modeling training procedure is used to pretrain the language models, as described in Section \ref{sec:bert}. Tokens are
masked at random with a uniform probability of 0.2. Input sequences are padded to the maximum length allowed by the model and truncated to the maximum
length, if greater. The maximum sequence length for BERT and RoBERTa models is 512 tokens \cite{devlin_bert_2019,liu_roberta_2019}. Based on the data
analysis in Section \ref{sec:data_analysis}, there are few, if any, samples that meet or exceed the maximum sequence length for BERT and RoBERTa
models.

@@

A train/test split of 70\% and 30\% and a validation split of 10\% of the training data was used for each model. The blank models (the models that are
trained from randomly initialized weights; also referred to as being trained from scratch) were trained for 200 epochs and the pretrained models were
trained for 100 epochs. These values were chosen through trial and error, monitoring the perplexity and loss curves of the
validation set to ensure that the pretraining does not result in net negative effects on the loss and perplexity of the model (i.e.~to ensure that the
models are not overfitting to the training data). Batch sizes of 16 were used for training all four models; this was the maximum batch size that
worked reliably with the hardware available to train the models\footnote{Values larger than 16 would sometimes cause CUDA Out of Memory errors, so 16
    was used to ensure all models trained continuously without error and/or interruption.}.

\begin{table}
    \centering
    \begin{tabular}{l r r}
        \toprule
        Corpus split & Split size (\%) & Samples \\
        \midrule
        Train        & 68.99           & 34,470  \\
        Validation   & 6.999           & 3,830   \\
        Test         & 30.01           & 16,415  \\
        \bottomrule
    \end{tabular}
    \caption{Number of samples and percentage of overall corpus by split.}
    \label{tab:corpus_splits}
\end{table}

\subsection{Performance Metrics}\label{sec:performance_metrics}
\textbf{Cross-Entropy Loss}. As mentioned previously, perplexity and cross-entropy loss (often shortened to loss) are used to track the performance of
the model on the training and validation sets, per epoch, while training. Cross-entropy loss is defined by the following equation:

\begin{equation}\label{eq:cross_entropy_loss}
    \begin{gathered}
        \mathcal{L}(x, y) = L = \{\mbox{$l_1$, $l_2$, ..., $l_N$}\}^T\\
        l_n = -\log \left[\frac{\exp(x_{n,y_n})}{\sum_{c=1}^{C}\exp(x_{n,c})}\right] \quad \forall n=1..N
    \end{gathered}
\end{equation}

\noindent
where $x$ is a $512 \times N$ vector representing model inputs (a sequence of tokens) and $y$ is a $1 \times N$ vector representing the target/label
for each sample in the batch. $L$ is an $N \times 1$ vector containing the losses for each sample in the batch, $N$ is the number of samples in the
batch, $l_n$ is the loss for an individual sample, $n$, in the batch, and $C$ is the number of class labels. For the purpose of language modeling, the
set of class labels is equivalent to the vocabulary of the tokenizer (and thus the language model). Numerical labels will correspond to token IDs. The
overall loss is then found by taking the mean of the loss over the batch, likewise for an epoch, the loss is found by taking the average of each batch
in the epoch.

\textbf{Perplexity}. Perplexity is related to the probability of a token appearing in a sequence. It is used as a measure of how well a language model
can predict tokens in a sequence and is defined by the equation below:

\begin{equation}\label{eq:perplexity}
    \begin{gathered}
        X = \{x_1, x_2, ..., x_N\}\\
        \mbox{PPL} = \exp\left[\frac{\sum_{i=1}^N - \log_2 x_i}{N}\right]
    \end{gathered}
\end{equation}

\noindent
Where $X$ is an input sequence, $N$ is the length of the sequence, and $x_i$ is an individual token in the sequence, $X$.

>>>>

\section{Data Sources}\label{sec:data_source}
There are four corpora used in this thesis. The datasets are primarily speech corpora intended either for automatic speech recognition research in the
domain of aviation communications, linguistics research in aviation English, or both. All data in all four corpora include both recordings and
transcriptions, which are audio and text, respectively. The transcriptions are used to pretrain and train the language models and tokenizers
(described in sections \ref{sec:language_modeling} and \ref{sec:tokenizers}).

@@

All data was processed in Python using primarily built-in functions/modules\footnote{The \lstinline|re| and \lstinline|xml| modules were also used
    although these are built into most Python distributions by default.}. The text data is extracted from the corpus transcripts by parsing from the
corpora-specific formats where necessary i.e.~Lisp in ATCC, XML in ATCO2 and ZCU CZ ATC, and plain text in ATCOSIM, the transcriber annotations
are removed where necessary\footnote{A copy of the original data is made, then the annotations are removed from the copy.}, and the resulting text
is written to a file in the standard corpus format (an ASCII text file with one line in the file corresponding to one sample from the corpus).

\subsection{Handling Spelling Errors \& Inconsistencies}\label{sec:spelling_errors}
By the nature of the data labeling process, there is a high likelihood to be spelling errors introduced by the annotators during corpora creation.
This gets exacerbated by the aggregation of several corpora with varying controls for correcting spelling errors and inconsistencies. Nearly all of
the errors are recoverable/correctable given context. For example, ``possibility of \textbf{tornado's}'' should clearly have been ``possibility of
\textbf{tornadoes}'', changing the possessive form of ``tornado'' to the plural form. There are also instances of spelling conventions that are
consistent within corpora, but become inconsistent when the corpora are combined.

@@

% hesitations, cut off words, different kinds of filled pauses
The mapping system described above is also used to modify within corpora conventions to be consistent across corpora. Three areas are addressed here:
hesitations, cut-off or partial words, and filled pauses within speech. These areas are defined differently between the four corpora, which makes
unifying the transcripts difficult. Hesitations, for example, are defined by one corpus as any incomplete or non-english speech sounds and left
completely undefined by another (despite annotators making notes about hesitations). The way in which hesitations, pauses, and partial words are
transcribed differs between corpora as well. Two corpora prescribe special tokens to these phenomena such as ``[hes]'', ``[hesitation]'', ``<pause>'',
or similar, while the others transcribe the data as it sounds (i.e., ``uh'', ``eh'', or ``er'' for filled pauses and ``cir-'' or ``cir+'' for
instances that were cut-off mid pronunciation). In order to facilitate the unification of the transcripts for all four corpora, the conventions for
transcribing these phenomena are redefined in Table \ref{tab:phenomena_definitions}.

@@

The ATCC and ZCU CZ ATC corpora make up a significant portion of the data, ATCC alone makes up more than half (approximately 54\%) of the combined
data, and ATCC and ZCU CZ ATC together make up approximately 81\% of the data. The sequence lengths of the samples are fairly consistent across
corpora with at most about an 18\% percent difference between sequence lengths (calculated using equation (\ref{eq:percent_difference}), below). The
number of unique tokens does not scale additively to the combined corpora, which is as expected since there will be an overlap of tokens between
corpora. Aviation English generally conforms to and follows ICAO phraseology standards, so we would expect that many terms would be
shared across corpora with the exception of geographic/location specific terms, such as city, landmark, airport, and airline names, etc.~that
typically vary across regions.

\begin{equation}\label{eq:percent_difference}
    \mbox{PD} = \frac{|V_1 - V_2|}{\frac{V_1 + V_2}{2}} \times 100
\end{equation}

@@

Sci-Kit Learn's \cite{pedregosa_scikit-learn_2011} implementation of the LOF algorithm was used with a value of \(K = 35\) for the K-Nearest Neighbors
algorithm and Minkowski distance for distance metrics/calculations (in this case, since \(p=2\), Euclidean distance is effectively used for distance
calculations). A total of 47 samples were detected as outliers and removed for analysis and are summarized in Table \ref{tab:outlier_stats}.

@@

Although all of the outliers have longer sequence lengths than the typical sequence lengths of the other samples in the data, going the through the
outlying samples manually and reviewing each reveals that, for the most part, the samples are typical albeit somewhat complex communications.
With the exception of one outlier, all of the samples labeled as outlying by the LOF algorithm were kept in the data. Examples of the outliers are
listed in Appendix A in Section \ref{sec:appendix_a}, including the sample removed.

\section{Language Modeling}\label{sec:language_modeling}
% bert, roberta
% n-gram
This section introduces the chosen language model architectures as well as their pretrained checkpoints. Two transformer-based architectures are used,
BERT and RoBERTa, as well as one statistical/machine learning-based approach: N-gram language modeling.

The transformers are pre-trained using the same masked language modeling (MLM) objective presented in Devlin et al.~\cite{devlin_bert_2019} to train
BERT for bidirectional encoding representation and likewise in Liu et al.~\cite{liu_roberta_2019} for training RoBERTa. Given the results of the
RoBERTa pretraining and downstream task training thereafter, it was decided to eliminate the next sentence prediction (NSP) task that was used to
pretrain BERT. MLM and NSP pretraining are explained more in-depth in Section \ref{sec:bert}, below.

@@

\textbf{Masked Language Modeling}. The ability of BERT to represent text in a bidirectional context is achieved through the Masked Language Modeling
(MLM) pretraining task~\cite{devlin_bert_2019}. MLM pretraining chooses a certain percentage of token positions at random as labels for prediction,
those tokens are then replaced with either a ``mask'' token, a random token, or the original token according to preconfigured probabilities (BERT used
80\% for the mask token, 10\% for random tokens, and 10\% for the original/unchanged token). During training, the masked sequences are used as input
to the model and the original is used as the prediction target. The model prediction is scored against the target using cross-entropy loss.

\textbf{Next Sentence Prediction}. While MLM conditions the model to represent tokens within sequences and predict the most probable token based on
its position in the sequence, NSP is meant to condition the model to understand relationships between sentences/sequences to further support
downstream tasks. Since BERTs training set consists primarily of literary works and excerpts from Wikipedia, the model simply predicts whether two
sentences appear together (one after the other) in the original text, making this a binary classification task as opposed to MLM which is a
multi-class classification task. Instead of a single sequence, the input consists of two sequences separated by a ``separation'' token and
beginning with a ``classification'' token, these are represented by \lstinline|[SEP]| and \lstinline|[CLS]|, respectively in Devlin et
al.~\cite{devlin_bert_2019}.

For symmetry between pretraining tasks, the special tokens specifying the beginning and end of input sequences are shared between tasks. For both MLM
and NSP objectives, input sequences always start with a classification token and end with a separation token (see Figure
\ref{fig:bert_mlm_input_example}). Likewise, NSP inputs always begin with a classification token and each input sequence is concatenated with a
separation token terminating each sequence (see Figure \ref{fig:bert_nsp_input_example} and equation (\ref{eq:bert_input_example})).

\begin{equation}\label{eq:bert_input_example}
    \begin{gathered}
        \mbox{[CLS], the, quick, brown, fox, [SEP]}\\
        \mbox{[CLS], the, quick, brown, fox, [SEP], jumps, over, the, [SEP]}
    \end{gathered}
\end{equation}

@@

@@

The pretraining methodology introduced by RoBERTa eliminated the NSP pretraining task that was originally used in BERT and opted instead to increase
the quantity of training data along with the batch sizes in addition to training the model longer. Lastly, there is a small difference in the
construction of input sequences as compared to BERT; input sequences always begin with a classification token (\lstinline|[CLS]|) and individual
segments are separated by a separation token (\lstinline|[SEP]|). Rather than having sequences end with a separation token as well, they are
terminated with an end of sequence token (\lstinline|[EOS]|). See equation (\ref{eq:roberta_input_example}) modified from Liu et
al.~\cite{liu_roberta_2019}, below.

@@

\section{Tokenization Algorithms}\label{sec:tokenizers}
% present and explain the tokenization algorithms used (possibly which models they correspond to)
Tokenization is the process of separating or segmenting words in sentences into their component ``tokens''. The tokenized version of a sentence is
referred to as a sequence. The rules by which the tokens are created or segmented depend entirely on the algorithm used. The simplest and most
intuitive way to do this is by creating tokens based on whitespace i.e., each word in the corpus is treated as a token\footnote{The Huggingface
    Tokenizer library implements this as the \lstinline|WordLevel| tokenizer.}. While this is simple and works in theory, it does not take into
account the fact that not every word in a language will occur in a corpus. The inevitable possibility for a tokenizer to see a word that was not in
the original training corpus is known as the out-of-vocabulary (OOV) problem~\cite{wu_googles_2016}. The OOV problem has been addressed by several
different algorithms. For brevity's sake, only the tokenization algorithms used by the models in this thesis are defined and explained.

@@

In order to map individual tokens to numerical values or IDs, an index of the words that appear in a corpus must be created. The numerical values for
each token are derived from this index. For the WordLevel algorithm, this process is referred to as the training the tokenizer model (see Algorithm
\ref{alg:wordlevel_training}).

@@

The tokenization of words with shared roots but different suffixes is shown in Table \ref{tab:wordlevel_tokenization_shared_roots_example}. All three
words are seen as independent and unique words by the tokenizer, since they are mapped to their own integer representations.

@@

\subsection{WordPiece}\label{sec:wordpiece}
% Wu et al. 2016; Schuster and Nakajima 2012
The WordPiece algorithm was developed by Schuster and Nakajima \cite{schuster_japanese_2012} as a word segmentation algorithm for language modeling in
Japanese and Korean voice search systems. It is further used by Wu et al.~\cite{wu_googles_2016} for neural machine translation tasks. The algorithm
demonstrated proficiency and increased performance for the models used in both tasks and effectively addresses the OOV problem.

One of the primary stipulations for this algorithm is effective and efficient handling of OOV words such that none are produced during normal
operation. To achieve this, during training, the vocabulary of the tokenizer is initialized to a basic set of characters; since WordPiece is designed
to be language agnostic, the initial vocabulary is specified as all basic unicode characters in addition to all ASCII characters.

@@

Table \ref{tab:wordpiece_tokenization_example}, below, shows an example of a WordPiece tokenizer model output on a sentence that was seen during the
training procedure (trained on the same data as the WordLevel tokenizer in Section \ref{sec:wordlevel}). Since this sentence was in the training data,
the segmentation is identical to that of the WordLevel tokenizer in Table \ref{tab:wordlevel_tokenization_example}, however, tokenization of similar
words (e.g.~words with shared roots or similar contractions) are handled differently, as in Table \ref{tab:wordpiece_tokenization_shared_roots_example}.
Since each word has a shared root word in the contractions (as well as an apostrophe separating the contraction), the tokenizer segments each word and
represents them as groups of small, more common tokens (thus with higher probabilities of occurring). Additionally, the WordPiece algorithm
effectively handles OOV words. For example, the OOV tokens in Table \ref{tab:wordpiece_unk_word_similar}, with a shared root and similar contractions
as those in Table \ref{tab:wordpiece_tokenization_shared_roots_example}, which appear in the training data. The base of the contraction, ``you'',
occurs in the training data as well as the apostrophe in the contraction, so the tokenizer segments those parts of the word as tokens. The last part
of the word, ``ll'', does not appear at all in the training data, so it is broken down into its component characters which are represented as ``l''
(beginning of a word) and ``\#\#l'' (segmented part of a word that should be concatenated to the previous token when they are recombined). Comparing
this to the output of the WordLevel tokenizer on the same word in Table \ref{tab:wordlevel_unk_word}, the WordPiece algorithm effectively addresses
the OOV problem for words similar to those in the trining corpus and, according to the output of the WordPiece tokenizer in Table
\ref{tab:wordpiece_unk_word}, for words unlike those in the training corpus.

@@

\subsection{Byte-Pair Encoding}\label{sec:bpe}
% Sennrich et al. 2016; Gage 1994
The Byte-Pair Encoding (BPE) tokenizer algorithm was adapted from a data compression algorithm by Phillip Gage in 1994~\cite{gage_feb94_1994}. The
data compression algorithm works by replacing common pairs of bytes in data with an unused byte to reduce the overall size of the data. The
tokenization algorithm works under the same principle, merging frequent pairs of characters or tokens into one token~\cite{sennrich_neural_2016}.

@@

When testing the BPE algorithm on words that did not appear in the training data, it is clear that the BPE algorithm effectively deals with and
processes OOV words (tables \ref{tab:bpe_unk_similar} and \ref{tab:bpe_unk}), even when no subunits of the word appear in the training data (Table
\ref{tab:bpe_unk}).

@@

\textbf{Byte-Level Byte-Pair Encoding}. This is a subset of the BPE algorithm that is also commonly used for NLP tasks such as neural machine
translation, language modeling, and generative tasks (GPT-2 uses a byte-level BPE tokenizer~\cite{radford_language_2019}). As Huggingface's
\lstinline|tokenizers| library implements it, this is a preprocessing step for the tokenizer model that maps all bytes in a string/sentence to their
own unique and visible character. Functionally, byte-level BPE is nearly identical to the standard BPE algorithm, described above, except that the
beginning of word symbol is actually rendered and represented by a visible character in the tokenized sequence (see Table
\ref{tab:byte_level_bpe_example}).

@@

\section{Discussion}\label{sec:discussion}
Considering the success of deep learning models with transformer-based architectures for language modeling and natural language processing (NLP)
tasks, more generally, in the conversational and literary English domains, it follows that transformer architectures should see widespread use in the
domain of aviation English for domain-specific NLP tasks. However, according to the literature, transformer-based language models have seen sparse
use. The most notable example of the use of a transformer neural network is the application of BERT for knowledge extraction from notice to airmen
(NOTAM) messages \cite{arnold_knowledge_2022}.

The expectation is that the language models trained from scratch will outperform the pretrained language models due to the nature of the differences
between literary English (the language the pretrained models were trained on) and aviation English. Additionally, the data being modeled are
orthographic transcriptions, which further extends the difference between the training corpora, thus leading to the hypothesis of the differences
in model performance. To test this hypothesis, the models will be divided into two sets or classes: (1) models trained from scratch and (2) pretrained
models that are fine-tuned (or transfer learned) using the data aggregated for this thesis (as described in Section \ref{sec:data_source}), these are
denoted as the sets of \textit{scratch} and \textit{pretrained} models, respectively. Each class includes a version of the BERT and RoBERTa models,
starting from scratch or the pretrained checkpoint, depending on the class of the model. In order to perform a fair comparison between models, the
following parameters (and/or hyperparameters) are kept the same across all four models:

\begin{itemize}
    \item Optimization algorithm
    \item Optimizer hyperparameters (see Table \ref{tab:optim_params})
    \item Number of hidden layers
    \item Number of neurons per layer
    \item Number of attention heads
    \item Activation functions
    \item Type of tokenizer algorithm
    \item Training, testing, and validation data
\end{itemize}

\subsection{Language Model Pretraining}\label{sec:lm_pretraining}
A standard masked language modeling training procedure is used to pretrain the language models, as described in Section \ref{sec:bert}. Tokens are
masked at random as described in the MLM process in Section \ref{sec:bert}. Input sequences are padded to the maximum length allowed by the model with
a special \lstinline|[PAD]| token and truncated to the maximum length, if greater. The maximum sequence length for BERT and RoBERTa models is 512
tokens \cite{devlin_bert_2019,liu_roberta_2019}. Based on the data analysis in Section \ref{sec:data_analysis}, there are few, if any, samples that
meet or exceed the maximum sequence length for BERT and RoBERTa models.

@@

A train/test split of 70\% and 30\% and a validation split of 10\% of the training data was used for each model. The models trained from scratch were
trained for 200 epochs and the pretrained models were trained for 100 epochs. These values were chosen through trial and error, monitoring the
perplexity and loss curves of the validation set to ensure that the pretraining does not result in net negative effects on the loss and perplexity of
the model (i.e.~to ensure that the models are not overfitting to the training data). Batch sizes of 16 were used for training all four models; this
was the maximum batch size that worked reliably with the hardware available to train the models\footnote{Values larger than 16 would sometimes cause
    Out of Memory errors on the GPU device used for training, so 16 was used to ensure all models trained continuously without error and/or
    interruption. Models were trained on a workstation 64 GB of RAM and an NVIDIA RTX 3090 GPU with 24 GB of VRAM.}.

\begin{table}[!t]
    \centering
    \begin{tabular}{l r r}
        \toprule
        Corpus split & Split size (\%) & Samples \\
        \midrule
        Train        & 68.99           & 34,470  \\
        Validation   & 6.999           & 3,830   \\
        Test         & 30.01           & 16,415  \\
        \bottomrule
    \end{tabular}
    \caption{Number of samples and percentage of overall corpus by split.}
    \label{tab:corpus_splits}
\end{table}

\subsection{Performance Metrics}\label{sec:performance_metrics}
\textbf{Cross-Entropy Loss}. As mentioned previously, perplexity and cross-entropy loss (often shortened to loss) are used to track the performance of
the model on the training and validation sets, per epoch, while training. Cross-entropy loss is defined below.

Let $\mathcal{L}(x, y)$ be the cross-entropy loss function that outputs a $N \times 1$ vector, $L$, that contains the loss values for each sample in a
batch of $N$ samples.

\begin{equation}\label{eq:cross_entropy_defs}
    \mathcal{L}(x, y) = L = \{\mbox{$l_1$, $l_2$, ..., $l_N$}\}^T
\end{equation}

\noindent
Where $x$ is a $512 \times N$ vector representing model inputs (a sequence of tokens) and $y$ is a $1 \times N$ vector representing the target/label
for each sample in the batch. The cross-entropy loss for a single sample is then calculated as

\begin{equation}\label{eq:cross_entropy_loss}
    l_n = -\log \left[\frac{\exp(x_{n,y_n})}{\sum_{c=1}^{C}\exp(x_{n,c})}\right] \quad \forall n=1..N
\end{equation}

\noindent
where $l_n$ is the loss for an individual sample in the batch, and $C$ is the number of class labels. For the purpose of language modeling, the
set of class labels is equivalent to the vocabulary of the tokenizer (and thus the language model). Numerical labels will correspond to token IDs. The
loss for a batch is calculated as the mean of the loss for each sample in the batch.

\begin{equation}\label{eq:cross_entropy_mean}
    \mathcal{L}(x, y) = \frac{1}{N}\sum_{i=1}^{N}l_i
\end{equation}

\noindent
Equation (\ref{eq:cross_entropy_mean}) can then be extended to find the mean cross-entropy loss over an epoch by swapping $l_i$ for batch means and
$\mathcal{L}(x, y)$ for the epoch mean. $N$ would then represent the number of batch in an epoch rather than the number of samples in a batch.

\textbf{Perplexity}. This metric is used to measure a models ability to predict the most probable token in a sequence. It will be used to measure and
track how well the language models can predict the most probable token for a specific point in a sequence.
